Spike spectra for recurrences

### 2022-07-21 Manuscript/inter_spike_manuscript_mdpi.tex
tiny changes according to MAs comments

Recurrence Plots (RPs) provide a vivid representation of complex dynamics stemming from potentially high dimensional systems, Eq.~\eqref{eq_rp_definition}. 
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N].
\end{equation}
The simple idea to 
track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\begin{figure}
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_spectrum_example}
 \caption{Schematic illustration of a $\tau$-recurrence rate based spectrum. \textbf{A} $x$-component time series of the Lorenz63-System (Eq.~\eqref{eq_model_Lorenz63}) and 
 \textbf{B} its corresponding Fourier power spectrum. 
 \textbf{C} Reconstructed state space portrait from the time series shown in \textbf{A} using PECUZAL time-delay embedding \cite{Kraemer2021}. 
 \textbf{D} Subset of the recurrence plot and the corresponding $\tau$-recurrence rate obtained from the state space trajectory in \textbf{C}. \sout{The shaded interval in the time series in \textbf{A} corresponds to the shown subset.} 
 \textbf{E} Fourier Power spectrum obtained from the $\tau$-recurrence (subset shown in panel \textbf{D}) \cite{Zbilut2008}. \todo[inline]{\textbf{D} and \textbf{E} show the results of a part of the time series in \textbf{A} which is highlighted in pink. }\todo[inline]{I think Fig.~\ref{fig_tau_rr_spectrum_example} should be restructured. We could talk about it.}
 }\label{fig_tau_rr_spectrum_example}
\end{figure}

@@

We consider the Logistic map $x_{n+1}=r\cdot x_n \left( 1-x_n \right)$ for changing control parameter $r$. We vary $r$ from $r=3.4$ to $r=4$ in steps of $0.001$. For 
each setting of $r$ 
\begin{itemize}[noitemsep]
\item[(1)] a time series of length $N=201$ is computed with a random initial condition $u_0 \in [0,\ 1]$, neglecting the first $1,000$ samples as transients,
\item[(2)] $100$ iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000} are computed,
\item[(3)] the time series and its iAAFT surrogates are embedded in a 2-dimensional state space using a time delay of unity,
\item[(4)] from the 2-dimensional trajectories RPs, Eq.~\eqref{eq_rp_definition}, are computed under a threshold $\varepsilon=0.05$,
\item[(5)] $\tau$-RR, Eq.~\eqref{eq_tau_rr}, is computed from the RP of the signal and from the RPs of the surrogates,
\item[(6)] inter spike spectra are obtained from $\tau$-RR of the signal and from the $\tau$-RRs of the surrogates, see Section \ref{sec_tau_rr_method}, and finally,
\item[(7)] from the distribution of the surrogate inter spike spectra the $95^\text{th}$ percentile is computed. The peaks of the inter spike spectrum of the signal which exceed 
this percentile are counted. 
\end{itemize}
In this example, the according null hypothesis is that the data stems from a process which yields the same auto-correlation, hence the same Fourier powerspectrum, and the same 
amplitude distribution. We consider the number of significant peaks in the inter spike spectrum with
respect to the control parameter in order to distinguish the corresponding dynamics (Fig.~\ref{fig_tau_rr_logistic}C).
A correlation with the positive Lyapunov exponent (Fig.~\ref{fig_tau_rr_logistic}A) is discernible 
($\rho_{\text{Pearson}}=0.72$). Moreover, this analysis can tackle period-doubling, since it ``measures'' the dominant cycles via the inter spike spectrum.

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_tau_rr_logistic}
 \caption{\textbf{A} Lyapunov exponent of the Logistic map as a function of the control parameter $r$. 
 \textbf{B} Number of significant peaks ($\alpha=0.05$) in the inter spike spectrum of the $\tau$-RR and its Pearson correlation coefficient to the Lyapunov exponent shown in \textbf{A} 
 (white noise surrogates). 
 \textbf{C} Same as \textbf{B}, but for iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000}. For obtaining the inter spike spectra we used 
 a LASSO regression and a regularization threshold corresponding to $\rho=0.95$ accordance of $\tau$-RR's and re-composed signals.  
}
\label{fig_tau_rr_logistic}
\end{figure}

@@

Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

>>>>

Recurrence Plots (RPs) provide a vivid representation of complex dynamics stemming from potentially high dimensional systems, Eq.~\eqref{eq_rp_definition}. 
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N].
\end{equation}
The crucial free parameter $\varepsilon$ (recurrence threshold) determines the visible structures in the RP. It can be chosen such, that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\begin{figure}
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_spectrum_example}
 \caption{Schematic illustration of a $\tau$-recurrence rate based spectrum. \textbf{A} $x$-component time series of the Lorenz63-System (Eq.~\eqref{eq_model_Lorenz63}) and 
 \textbf{B} its corresponding Fourier power spectrum. 
 \textbf{C} Reconstructed state space portrait from the time series shown in \textbf{A} using PECUZAL time-delay embedding \cite{Kraemer2021}. 
 \textbf{D} Subset of the recurrence plot and the corresponding $\tau$-recurrence rate obtained from the state space trajectory in \textbf{C}. 
 \textbf{E} Fourier Power spectrum obtained from the $\tau$-recurrence rate (subset shown in panel \textbf{D}) \cite{Zbilut2008}. \textbf{D} and \textbf{E} show the results of a part of the time series, 
 which is highlighted in pink in \textbf{A}.\todo[inline]{MA: I think Fig.~\ref{fig_tau_rr_spectrum_example} should be restructured. We could talk about it.}
 }\label{fig_tau_rr_spectrum_example}
\end{figure}

@@

We consider the Logistic map $x_{n+1}=r\cdot x_n \left( 1-x_n \right)$ for changing control parameter $r$. We vary $r$ from $r=3.4$ to $r=4$ in steps of $0.001$. For 
each setting of $r$ 
\begin{itemize}[noitemsep]
\item[(1)] a time series of length $N=201$ is computed with a random initial condition $u_0 \in [0,\ 1]$, neglecting the first $1,000$ samples as transients,
\item[(2)] $100$ iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000} are computed,
\item[(3)] the time series and its iAAFT surrogates are embedded in a 2-dimensional state space using a time delay of unity,
\item[(4)] from the 2-dimensional trajectories RPs, Eq.~\eqref{eq_rp_definition}, are computed under a threshold $\varepsilon=0.05$,
\item[(5)] $\tau$-RR, Eq.~\eqref{eq_tau_rr}, is computed from the RP of the signal and from the RPs of the surrogates,
\item[(6)] inter spike spectra are obtained from $\tau$-RR of the signal and from the $\tau$-RRs of the surrogates, see Section \ref{sec_tau_rr_method}, and finally,
\item[(7)] from the distribution of the surrogate inter spike spectra the $95^\text{th}$ percentile is computed. The peaks of the inter spike spectrum of the signal which exceed 
this percentile are counted. 
\end{itemize}
In this example, the according null hypothesis for constructing the surrogate data is that the data stems from a process which yields the same auto-correlation, 
hence the same Fourier powerspectrum, and the same 
amplitude distribution. We consider the number of significant peaks in the inter spike spectrum with
respect to the control parameter in order to distinguish the corresponding dynamics (Fig.~\ref{fig_tau_rr_logistic}C).
A correlation with the positive Lyapunov exponent (Fig.~\ref{fig_tau_rr_logistic}A) is discernible 
($\rho_{\text{Pearson}}=0.72$). Moreover, this analysis can tackle period-doubling, since it ``measures'' the dominant cycles via the inter spike spectrum. \todo{MA: This part is not understandable}

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_tau_rr_logistic}
 \caption{\textbf{A} Bifurcation diagram and Lyapunov exponent and of the Logistic map as a function of the control parameter $r$.
 \textbf{B} Number of significant peaks ($\alpha=0.05$) in the inter spike spectrum of the $\tau$-RR and its Pearson correlation coefficient to the Lyapunov exponent shown in \textbf{A} 
 (white noise surrogates). 
 \textbf{C} Same as \textbf{B}, but for iterative Amplitude Adjusted Fourier Transform (iAAFT) surrogates \cite{Schreiber1996,Schreiber2000}. For obtaining the inter spike spectra we used 
 a LASSO regression and a regularization threshold corresponding to $\rho=0.95$ accordance of $\tau$-RR's and re-composed signals.  
}
\label{fig_tau_rr_logistic}
\end{figure}

@@

\begin{itemize}
\item computationally expensive.
\item STLS and LASSO yield different results.
\item incorporates FFT-peaks (see Vostok data).
\item bifurcation detection only possible when additional period(s) is not an integer multiple of the former period(s).
\end{itemize}
